---
title: "01_dataset_creation"
author: "Claudia Jiménez"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
# output:
#   html_notebook:
#     code_folding: hide
#     number_sections: yes
#     theme: flatly
#     toc: yes
#     toc_float: yes
---

# Environment

```{r message=FALSE, warning=FALSE}
library(bigrquery)
library(DBI)
library(dplyr)
library(forcats)
library(summarytools)
library(ggplot2)
library(DiagrammeR)
```

# BigQuery related functions

This chunks creates the run_query and getSQL function.

```{r}
# Function that takes in a SQL command and runs it on BigQuery. This avoids the connection data in each iteration
run_query<-function(query){
  query_output<-dbGetQuery(con,query)
  return(query_output)
}

# Function for reading SQL files from a folder
getSQL <- function(filepath) {
  con = file(filepath, "r")
  sql.string <- ""
  while (TRUE) {
    line <- readLines(con, n = 1)
    if (length(line) == 0) {
      break
    }
    line <- gsub("\\t", " ", line)
    if (grepl("--", line) == TRUE) {
      line <- paste(sub("--", "/*", line), "*/")
    }
    sql.string <- paste(sql.string, line)
  }
  close(con)
  return(sql.string)
}
```

## Setting up connection

```{r}
bigrquery::bq_auth() # UNCOMMENT WHEN HAVEN TOCKEN ISSUES!!!
# Establecer la conexión a BigQuery
project_id <- readline(prompt = "Enter your project ID: ")

con <- dbConnect(
  bigrquery::bigquery(),
  project = project_id
)

```


# Data Extraction

Loading queries and extracting the data.

```{r}
## Demographics dates
basic_demographics <- run_query(getSQL("sql/basic_demographics.sql"))
icu_mortality_72hrs <- run_query(getSQL("sql/icu_mortality_72hrs.sql"))
apachescore <- run_query(getSQL("sql/apachescore.sql"))
apacheadmission <- run_query(getSQL("sql/apacheadmission.sql"))
prev_dept_los <- run_query(getSQL("sql/prev_dept_los.sql"))
clearance_pivoted_vitals <- run_query(getSQL("sql/clearance_pivoted_vital.sql"))
clearance_pivoted_lab <- run_query(getSQL("sql/clearance_pivoted_lab.sql"))
pivoted_med <- run_query(getSQL("sql/pivoted_med.sql"))
```

#Imputation: 1º part
##Imputation dataframe clearance_pivoted_vitals
```{r}
#Replace NA with median: clearance_pivoted_vitals. Values firts and last
clearance_piv_vital <- clearance_pivoted_vitals %>%
  mutate(
    heartrate_first = ifelse(is.na(heartrate_first), median(heartrate_first, na.rm = TRUE), heartrate_first),
    heartrate_last = ifelse(is.na(heartrate_last), median(heartrate_last, na.rm = TRUE), heartrate_last),
    respiratoryrate_first = ifelse(is.na(respiratoryrate_first), median(respiratoryrate_first, na.rm = TRUE), respiratoryrate_first),
    respiratoryrate_last = ifelse(is.na(respiratoryrate_last), median(respiratoryrate_last, na.rm = TRUE), respiratoryrate_last),
    spo2_first = ifelse(is.na(spo2_first), median(spo2_first, na.rm = TRUE), spo2_first),
    spo2_last = ifelse(is.na(spo2_last), median(spo2_last, na.rm = TRUE), spo2_last),
    nibp_systolic_first = ifelse(is.na(nibp_systolic_first), median(nibp_systolic_first, na.rm = TRUE), nibp_systolic_first),
    nibp_systolic_last = ifelse(is.na(nibp_systolic_last), median(nibp_systolic_last, na.rm = TRUE), nibp_systolic_last),
    nibp_diastolic_first = ifelse(is.na(nibp_diastolic_first), median(nibp_diastolic_first, na.rm = TRUE), nibp_diastolic_first),
    nibp_diastolic_last = ifelse(is.na(nibp_diastolic_last), median(nibp_diastolic_last, na.rm = TRUE), nibp_diastolic_last),
    nibp_mean_first = ifelse(is.na(nibp_mean_first), median(nibp_mean_first, na.rm = TRUE), nibp_mean_first),
    nibp_mean_last = ifelse(is.na(nibp_mean_last), median(nibp_mean_last, na.rm = TRUE), nibp_mean_last),
    temperature_first = ifelse(is.na(temperature_first), median(temperature_first, na.rm = TRUE), temperature_first),
    temperature_last = ifelse(is.na(temperature_last), median(temperature_last, na.rm = TRUE), temperature_last),
    ibp_systolic_first = ifelse(is.na(ibp_systolic_first), median(ibp_systolic_first, na.rm = TRUE), ibp_systolic_first),
    ibp_systolic_last = ifelse(is.na(ibp_systolic_last), median(ibp_systolic_last, na.rm = TRUE), ibp_systolic_last),
    ibp_diastolic_first = ifelse(is.na(ibp_diastolic_first), median(ibp_diastolic_first, na.rm = TRUE), ibp_diastolic_first),
    ibp_diastolic_last = ifelse(is.na(ibp_diastolic_last), median(ibp_diastolic_last, na.rm = TRUE), ibp_diastolic_last),
    ibp_mean_first = ifelse(is.na(ibp_mean_first), median(ibp_mean_first, na.rm = TRUE), ibp_mean_first),
    ibp_mean_last = ifelse(is.na(ibp_mean_last), median(ibp_mean_last, na.rm = TRUE), ibp_mean_last)
  )

#Replace NA with difference: clearance_pivoted_vitals. Values clear
clearance_piv_vital <- clearance_piv_vital %>%
  mutate(
    heartrate_clear = ifelse(is.na(heartrate_clear), heartrate_last - heartrate_first, heartrate_clear),
    respiratoryrate_clear = ifelse(is.na(respiratoryrate_clear), respiratoryrate_last - respiratoryrate_first, respiratoryrate_clear),
    spo2_clear = ifelse(is.na(spo2_clear), spo2_last - spo2_first, spo2_clear),
    nibp_systolic_clear = ifelse(is.na(nibp_systolic_clear), nibp_systolic_last - nibp_systolic_first, nibp_systolic_clear),
    nibp_diastolic_clear = ifelse(is.na(nibp_diastolic_clear), nibp_diastolic_last - nibp_diastolic_first, nibp_diastolic_clear),
    nibp_mean_clear = ifelse(is.na(nibp_mean_clear), nibp_mean_last - nibp_mean_first, nibp_mean_clear),
    temperature_clear = ifelse(is.na(temperature_clear), temperature_last - temperature_first, temperature_clear),
    ibp_systolic_clear = ifelse(is.na(ibp_systolic_clear), ibp_systolic_last - ibp_systolic_first, ibp_systolic_clear),
    ibp_diastolic_clear = ifelse(is.na(ibp_diastolic_clear), ibp_diastolic_last - ibp_diastolic_first, ibp_diastolic_clear),
    ibp_mean_clear = ifelse(is.na(ibp_mean_clear), ibp_mean_last - ibp_mean_first, ibp_mean_clear)
  )


```
##Imputation dataframe clearance_pivoted_lab
```{r}
clearance_piv_lab <- clearance_pivoted_lab %>%
  mutate(
    albumin_first = ifelse(is.na(albumin_first), median(albumin_first, na.rm = TRUE), albumin_first),
    albumin_last = ifelse(is.na(albumin_last), median(albumin_last, na.rm = TRUE), albumin_last),
    bilirubin_first = ifelse(is.na(bilirubin_first), median(bilirubin_first, na.rm = TRUE), bilirubin_first),
    bilirubin_last = ifelse(is.na(bilirubin_last), median(bilirubin_last, na.rm = TRUE), bilirubin_last),
    BUN_first = ifelse(is.na(BUN_first), median(BUN_first, na.rm = TRUE), BUN_first),
    BUN_last = ifelse(is.na(BUN_last), median(BUN_last, na.rm = TRUE), BUN_last),
    calcium_first = ifelse(is.na(calcium_first), median(calcium_first, na.rm = TRUE), calcium_first),
    calcium_last = ifelse(is.na(calcium_last), median(calcium_last, na.rm = TRUE), calcium_last),
    creatinine_first = ifelse(is.na(creatinine_first), median(creatinine_first, na.rm = TRUE), creatinine_first),
    creatinine_last = ifelse(is.na(creatinine_last), median(creatinine_last, na.rm = TRUE), creatinine_last),
    glucose_first = ifelse(is.na(glucose_first), median(glucose_first, na.rm = TRUE), glucose_first),
    glucose_last = ifelse(is.na(glucose_last), median(glucose_last, na.rm = TRUE), glucose_last),
    bicarbonate_first = ifelse(is.na(bicarbonate_first), median(bicarbonate_first, na.rm = TRUE), bicarbonate_first),
    bicarbonate_last = ifelse(is.na(bicarbonate_last), median(bicarbonate_last, na.rm = TRUE), bicarbonate_last),
    TotalCO2_first = ifelse(is.na(TotalCO2_first), median(TotalCO2_first, na.rm = TRUE), TotalCO2_first),
    TotalCO2_last = ifelse(is.na(TotalCO2_last), median(TotalCO2_last, na.rm = TRUE), TotalCO2_last),
    hematocrit_first = ifelse(is.na(hematocrit_first), median(hematocrit_first, na.rm = TRUE), hematocrit_first),
    hematocrit_last = ifelse(is.na(hematocrit_last), median(hematocrit_last, na.rm = TRUE), hematocrit_last),
    hemoglobin_first = ifelse(is.na(hemoglobin_first), median(hemoglobin_first, na.rm = TRUE), hemoglobin_first),
    hemoglobin_last = ifelse(is.na(hemoglobin_last), median(hemoglobin_last, na.rm = TRUE), hemoglobin_last),
    INR_first = ifelse(is.na(INR_first), median(INR_first, na.rm = TRUE), INR_first),
    INR_last = ifelse(is.na(INR_last), median(INR_last, na.rm = TRUE), INR_last),
    lactate_first = ifelse(is.na(lactate_first), median(lactate_first, na.rm = TRUE), lactate_first),
    lactate_last = ifelse(is.na(lactate_last), median(lactate_last, na.rm = TRUE), lactate_last),
    platelets_first = ifelse(is.na(platelets_first), median(platelets_first, na.rm = TRUE), platelets_first),
    platelets_last = ifelse(is.na(platelets_last), median(platelets_last, na.rm = TRUE), platelets_last),
    potassium_first = ifelse(is.na(potassium_first), median(potassium_first, na.rm = TRUE), potassium_first),
    potassium_last = ifelse(is.na(potassium_last), median(potassium_last, na.rm = TRUE), potassium_last),
    ptt_first = ifelse(is.na(ptt_first), median(ptt_first, na.rm = TRUE), ptt_first),
    ptt_last = ifelse(is.na(ptt_last), median(ptt_last, na.rm = TRUE), ptt_last),
    sodium_first = ifelse(is.na(sodium_first), median(sodium_first, na.rm = TRUE), sodium_first),
    sodium_last = ifelse(is.na(sodium_last), median(sodium_last, na.rm = TRUE), sodium_last),
    wbc_first = ifelse(is.na(wbc_first), median(wbc_first, na.rm = TRUE), wbc_first),
    wbc_last = ifelse(is.na(wbc_last), median(wbc_last, na.rm = TRUE), wbc_last),
    bands_first = ifelse(is.na(bands_first), median(bands_first, na.rm = TRUE), bands_first),
    bands_last = ifelse(is.na(bands_last), median(bands_last, na.rm = TRUE), bands_last),
    alt_first = ifelse(is.na(alt_first), median(alt_first, na.rm = TRUE), alt_first),
    alt_last = ifelse(is.na(alt_last), median(alt_last, na.rm = TRUE), alt_last),
    ast_first = ifelse(is.na(ast_first), median(ast_first, na.rm = TRUE), ast_first),
    ast_last = ifelse(is.na(ast_last), median(ast_last, na.rm = TRUE), ast_last),
    alp_first = ifelse(is.na(alp_first), median(alp_first, na.rm = TRUE), alp_first),
    alp_last = ifelse(is.na(alp_last), median(alp_last, na.rm = TRUE), alp_last)
  )

clearance_piv_lab <- clearance_piv_lab %>%
  mutate(
    albumin_clear = ifelse(is.na(albumin_clear), albumin_last - albumin_first, albumin_clear),
    bilirubin_clear = ifelse(is.na(bilirubin_clear), bilirubin_last - bilirubin_first, bilirubin_clear),
    BUN_clear = ifelse(is.na(BUN_clear), BUN_last - BUN_first, BUN_clear),
    calcium_clear = ifelse(is.na(calcium_clear), calcium_last - calcium_first, calcium_clear),
    creatinine_clear = ifelse(is.na(creatinine_clear), creatinine_last - creatinine_first, creatinine_clear),
    glucose_clear = ifelse(is.na(glucose_clear), glucose_last - glucose_first, glucose_clear),
    bicarbonate_clear = ifelse(is.na(bicarbonate_clear), bicarbonate_last - bicarbonate_first, bicarbonate_clear),
    TotalCO2_clear = ifelse(is.na(TotalCO2_clear), TotalCO2_last - TotalCO2_first, TotalCO2_clear),
    hematocrit_clear = ifelse(is.na(hematocrit_clear), hematocrit_last - hematocrit_first, hematocrit_clear),
    hemoglobin_clear = ifelse(is.na(hemoglobin_clear), hemoglobin_last - hemoglobin_first, hemoglobin_clear),
    INR_clear = ifelse(is.na(INR_clear), INR_last - INR_first, INR_clear),
    lactate_clear = ifelse(is.na(lactate_clear), lactate_last - lactate_first, lactate_clear),
    platelets_clear = ifelse(is.na(platelets_clear), platelets_last - platelets_first, platelets_clear),
    potassium_clear = ifelse(is.na(potassium_clear), potassium_last - potassium_first, potassium_clear),
    ptt_clear = ifelse(is.na(ptt_clear), ptt_last - ptt_first, ptt_clear),
    sodium_clear = ifelse(is.na(sodium_clear), sodium_last - sodium_first, sodium_clear),
    wbc_clear = ifelse(is.na(wbc_clear), wbc_last - wbc_first, wbc_clear),
    bands_clear = ifelse(is.na(bands_clear), bands_last - bands_first, bands_clear),
    alt_clear = ifelse(is.na(alt_clear), alt_last - alt_first, alt_clear),
    ast_clear = ifelse(is.na(ast_clear), ast_last - ast_first, ast_clear),
    alp_clear = ifelse(is.na(alp_clear), alp_last - alp_first, alp_clear)
  )

```

# Joining dataframes

```{r}
# Define the data frames to join
dfs_to_join <- list(basic_demographics, clearance_piv_vital, icu_mortality_72hrs,apachescore, apacheadmission, prev_dept_los, clearance_piv_lab, pivoted_med)

# Perform the left join
final_df <- Reduce(function(x, y) merge(x, y, by = "patientunitstayid", all.x = TRUE), dfs_to_join)

```

## Creation of new variables and adjustment

```{r}
## Pass the variable character age to numeric. Replace >89 with 90.Rename variable gender (1 male, 2 female)

final_df <- final_df %>%
  mutate(
    age = ifelse(age == '> 89', 90,
                          ifelse(trimws(age) == '', NA_real_,
                                  as.numeric(age)))
  ) %>%
  mutate(
    sex = ifelse(
      sex == 2, "female",
      ifelse(
        sex == 1, "male", NA_character_
      )
    ) 
  ) %>%
  mutate(
    icu_mortality_72hrs = ifelse(
      icu_mortality_72hrs == 1, "Expired",
      ifelse(
        icu_mortality_72hrs == 0, "Alived", NA_character_
      )
    )
  )


```
#Imputation 2º part
```{r}
#Delete NA values of variable gender (177)
final_df <- final_df %>%
 mutate(
   sex = as.factor(fct_explicit_na(sex, na_level = "Unknown"))) #NA values of gender are converted to "Unknown"
#change reference level of factor gender variable to "Male") 
final_df$sex <- relevel(final_df$sex,ref = "male")
# Filtrar las filas donde gender no sea "Unknown"
final_df <- final_df %>% filter(sex != "Unknown")

#Delete NA value of variable icu_mortality_72hrs
final_df <- final_df %>%
  mutate(icu_mortality_72hrs = coalesce(icu_mortality_72hrs, "Alived"))

#Delete NA values of variable age (6)
final_df <- final_df %>%
  filter(!is.na(age))

# Change NA value of variable apachescoreIV = median
final_df <- final_df %>%
  mutate(apachescoreIV = ifelse(is.na(apachescoreIV), median(apachescoreIV, na.rm = TRUE), apachescoreIV))

#Change minutes to hours: variable prev_dept_los
final_df <- final_df %>%
  mutate(prev_dept_los_hrs = abs(prev_dept_los/60)) %>%
  select(-prev_dept_los)

```

##Imputation NA variable clear
```{r}
#Replace NA with 0 df:pivoted_med
final_df <- final_df %>%
mutate(
norepinephrine = coalesce(norepinephrine, 0),
epinephrine = coalesce(epinephrine, 0),
dopamine = coalesce(dopamine, 0),
dobutamine = coalesce(dobutamine, 0),
phenylephrine = coalesce(phenylephrine, 0),
vasopressin = coalesce(vasopressin, 0),
milrinone = coalesce(milrinone, 0),
heparin = coalesce(heparin, 0),
warfarin = coalesce(warfarin, 0),
)
```


# Order variables
```{r}
final_df_order <- final_df %>%
  select(patientunitstayid,age,sex,icu_mortality_72hrs,apachescoreIV,apachedxgroup,prev_dept_los_hrs,prev_dept,albumin_clear,bilirubin_clear,BUN_clear,calcium_clear,creatinine_clear,glucose_clear,bicarbonate_clear,TotalCO2_clear,hematocrit_clear,hemoglobin_clear,INR_clear,lactate_clear,platelets_clear,potassium_clear,ptt_clear,sodium_clear,wbc_clear,bands_clear,alt_clear,ast_clear,alp_clear,heartrate_clear,respiratoryrate_clear,spo2_clear,nibp_systolic_clear,nibp_diastolic_clear,nibp_mean_clear,temperature_clear,ibp_systolic_clear,ibp_diastolic_clear,ibp_mean_clear,norepinephrine,epinephrine,dopamine,dobutamine,phenylephrine,vasopressin,milrinone,heparin,warfarin
 )
```

# Exclusion criteria

```{r}
print('Initial number of patients')
a <- nrow(final_df)
a

final_df_order <- final_df %>% filter(age > 16)
print('Underage patients excluded')
b <- nrow(final_df_order)
a-b


print('Número final de pacientes')
c

data <- list(a,b,c)
grViz("digraph graph2 {
      graph [layout = dot]
      node [shape = rectanglewidth = 4, fillcolor = Beige]
      a[label = '@@1']
      b[label = '@@2]
      c[label = @@3]
      a -> a-b -> c
}
[1]:  paste0('Initial number of patients (n = ', final_df, ')')
[2]: paste0('Underage patients excluded (n = ', final_df_order, ')')
[3]: paste0('Número final de pacientes (n = ', Número final de pacientes, ')')
")

```
#Creation of exclusion criteria diagram
```{r}
# Crear un gráfico de flujo
gr <- create_graph()

# Añadir nodos al gráfico
gr %>%
  add_node("Inicio", shape = "ellipse") %>%
  add_node("Imprimir número inicial de pacientes", shape = "rectangle") %>%
  add_node("Obtener el número inicial de pacientes (a)", shape = "diamond") %>%
  add_node("Excluir pacientes menores de 16 años", shape = "rectangle") %>%
  add_node("Imprimir 'Pacientes menores de 16 años excluidos'", shape = "rectangle") %>%
  add_node("Obtener el número final de pacientes (b)", shape = "diamond") %>%
  add_node("Calcular la diferencia a - b", shape = "rectangle") %>%
  add_node("Imprimir 'Número final de pacientes'", shape = "rectangle") %>%
  add_node("Fin", shape = "ellipse")

# Añadir bordes (flechas) entre nodos
gr %>%
  add_edge("Inicio", "Imprimir número inicial de pacientes") %>%
  add_edge("Imprimir número inicial de pacientes", "Obtener el número inicial de pacientes (a)") %>%
  add_edge("Obtener el número inicial de pacientes (a)", "Excluir pacientes menores de 16 años") %>%
  add_edge("Excluir pacientes menores de 16 años", "Imprimir 'Pacientes menores de 16 años excluidos'") %>%
  add_edge("Imprimir 'Pacientes menores de 16 años excluidos'", "Obtener el número final de pacientes (b)") %>%
  add_edge("Obtener el número final de pacientes (b)", "Calcular la diferencia a - b") %>%
  add_edge("Calcular la diferencia a - b", "Imprimir 'Número final de pacientes'") %>%
  add_edge("Imprimir 'Número final de pacientes'", "Fin")

# Visualizar el gráfico
render_graph(gr)

```